Exploring dimensions of biodiversity in Japanese ferns

### 2020-08-25 ms/manuscript.Rmd
Update MS

loadd(list = c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp"
),
cache = ja_fern_cache)

@@

A necessary step in any analysis of biodiversity is to set the grain size used to define communities. This decision must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling. We tested using 0.1, 0.2, 0.3, and 0.4 degree grid-cells for defining communities (co-occurring species). At each size, species occurrences were converted into a presence-absence community matrix (a species was considered present if at least one specimen was recorded in that grid cell). We calculated sampling redundance (1 - richness/number of specimens) [@Garcillan2003] to quantify adequacy of sampling. Preliminary analysis indicated that 0.2 degree grid-cells are optimal for our dataset (`r s_figure("grain-size")`). We filtered out any cells with redundancy < 0.1 (mostly cells along the coast with very low land area) and used this for subsequent analyses.

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan using LUCID software [@Ebihara2016b; @Ebihara2017; @Ebihara2019b]. Quantitative traits were measured on 10 randomly selected specimens per species, and qualitative traits determined by observing voucher specimens [@Ebihara2016b; @Ebihara2017]. All qualitative traits were measured in binary format; \eg, a single trait with three states was formatted as three binary traits. Quantitative traits were measured in fertile (i.e., spore-bearing) and sterile fronds separately originally, then averaged. From the original trait list, we selected only traits with putatative ecological function, and removed any that had Pearson correlation coefficient >0.6 or fewer than three observations of a given trait state (`r table("traits-used")`).

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils availabsle for molecular dating. Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b with a globally sampled plastid dataset including all fern sequences on GenBank for four plastid genes (*rbcL*, *atpB*, *aptA*, and *rps4*) and 66 other coding, single-copy plastid genes extracted from 100 complete fern plastomes (Nitta et al., in prep) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total). We conducted phylogenetic analysis with IQTREE [@Nguyen2015] and molecular dating with treePL [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. We then trimmed the tree to Japanese taxa only and used this for all further analyses.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's lambda [@Pagel1999]. *K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under Brownian Motion (BM); lambda is a scaling parameter used to transform the phylogeny such that trait values fit those most closely expected under BM. Both lambda and *K* = 1 under BM; lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering). We measured *K* and lambda using the \function{phylosig} function in the R package \package{phytools} [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for lambda. We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the \function{phylo.d} function in the R package \package{caper} [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.

@@

We measured functional diversity (FD) following the method of @Petchey2002, which is a functional analog of PD. We first calculated distances using Gower's method [@Gower1971] on the trait data with the \function{gowdis} function in the R package \package{FD} [@Laliberte2011]. We weighted quantitative and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled quantitative traits prior to calculating distances. We then used the distances to build a trait dendrogram using the \function{hclust} function in the \package{stats} R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above.

@@

Observed values of PD, FD, and PE are known to be closely related to species richness: as more taxa are drawn at random without replacement from a given set, it becomes increasingly unlikely to draw a taxon that is distantly related (or functionally dissimilar) to the others. Therefore, we determined whether the observed metrics are more or less diverse than random for a given number of taxa by conducting a randomization test. We generated 999 random communities using the "independent swap" algorithm of @Gotelli2000, which conserves richness per site and species occurrence frequencies while randomizing species identities, then compared the observed value to the distribution of random values. Statistical significance was assessed using a two-tailed test with alpha = 0.5: observed values > 97.5% or < 0.25% were considered significant.

We tested whether endemic areas have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleo-endemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014]. Low RPE indicates short branchlengths (neoendemism) and high RPE indicates long branchlengths (paleoendemism). Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable to those calculated for the null distribution. We used the same set of null communities as for SES of PE, and carried out CANAPE using custom R scripts.

@@

We downloaded SHP files of protected areas in Japan from the Japan Ministry of the Environment (https://www.biodic.go.jp/biodiversity/activity/policy/map/map17/index.html). We only considered areas that include protection of plants, and categorized protected areas as either "high" (no human activities allowed at all) or "normal" (some human activities allowed by permit). We cropped grid cells to only include land area, then calculated the percent of protected area of each status type within grid cells with high levels of biodiversity as measured with PD, FD, PE, or taxonomic richness. For PD, FD, and PE, significance was assessed with a one-tailed test at alpha = 0.05 against the 999 null communities described above. For taxonomic richness, cells ranked in the top >5% were considered highly diverse.

@@

The topology of the phylogenetic tree was in general agreement with other recent, large plastid fern phylogenies. Ferns were recovered as monophyletic, as were all families and genera; taxonomy follows PPGI for the family level and the Japan Green List for the genus level and below. The age of the fern clade is estimated to be _ my.

The distribution of specimen collection locations is highly skewed, with >3,000 specimens per grid cell northwest of Tokyo in Saitama prefecture, and many fewer elsewhere (`r s_figure("abundance")`A). Despite this, sampling redundancy is generally high throughout the country, with areas of low redundancy in Hokkaido and some coastal areas and small islands (`r s_figure("abundance")`B). The grid cells that had to be excluded from the analysis due to low redundancy were mostly from Hokkaido.

@@

Results of phylogenetic signal analysis are given in Table _. All quantitative traits showed significant phylogenetic signal, but estimates of signal strength differed between K and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of K were < 0.06, indicating less phylogenetic signal than expected under BM. A minority of quantitative traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (D =< 0); others had D > 0.

@@

The latitudinal hump-shaped pattern of species richness observed in ferns here has been well-documented [@Kusumoto2017; @Ebihara2019b], and similar patterns have been observed in deciduous broadleaved trees and conifers [@Shiono2015]. This pattern is probably due at least in part to the distribution of land area in Japan: maximum land area is located at mid-latitudes, and area is related to species richness by the well-known species-area relationship [@Arrhenius1921]. There may also be a combined climatic and elevational effect: the southern part of Honshu is warmer and more humid than areas further north, while also harboring high elevational variation that may support a wide range of taxa, which is not present in the southern islands.

@@

>>>>

loadd(list = c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns"
),
cache = ja_fern_cache)

@@

A necessary step in any analysis of biodiversity is to set the grain size used to define communities. Here, one must consider that smaller grain size is needed to detect environmental effects, while larger grain size is needed to ensure adequate species sampling. We tested using 0.1°, 0.2°, 0.3°, and 0.4° grid-cells (size refers to length of side of the cell) for defining communities (co-occurring species). At each grain size, species occurrences were converted into a presence-absence community matrix (a species was considered present if at least one specimen was recorded in that grid cell). We calculated sampling redundance (1 - richness/number of specimens) [@Garcillan2003] to quantify adequacy of sampling. Preliminary analysis indicated that 0.2° grid-cells are optimal for our dataset (`r s_figure("grain-size")`). We filtered out any cells with redundancy < 0.1 and used this for subsequent analyses.

@@

We used traits originally compiled for identification of ferns and lycophytes of Japan as follows [@Ebihara2016b; @Ebihara2017]. Continuous traits were measured on 10 randomly selected specimens per species. Fertile (i.e., spore-bearing) and sterile fronds were measured separately, then averaged. Qualitative traits were scored by observing voucher specimens. All qualitative traits were scored in binary format; \eg, a single trait with three states was formatted as three binary traits. From the original trait list, we selected only traits with putatative ecological function (`r table("traits-used")`), and removed any that had Pearson correlation coefficient >0.6 or fewer than three observations of a given trait state.

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils available for molecular dating. Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b with a globally sampled plastid dataset including all fern sequences on GenBank for four plastid genes (*rbcL*, *atpB*, *aptA*, and *rps4*) and 66 other coding, single-copy plastid genes extracted from 100 complete fern plastomes (Nitta et al., in prep) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total). We conducted phylogenetic analysis with IQTREE [@Nguyen2015] and molecular dating with treePL [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. We then trimmed the tree to Japanese taxa only and used this for all further analyses.

@@

We tested for phylogenetic signal in continuous traits using two alternative metrics, Blomberg's *K* [@Blomberg2003] and Pagel's lambda [@Pagel1999]. *K* describes the ratio between the observed variance in a trait vs. the amount of variance expected under Brownian Motion (BM); lambda is a scaling parameter that transforms the phylogeny such that trait values fit those most closely expected under BM. Both lambda and *K* = 1 under BM; lower values indicate less signal (overdispersion) and higher values indicate more signal (clustering). We measured *K* and lambda using the \function{phylosig} function in the R package \package{phytools} [@Revell2012], which also calculates significance based on a randomization test for *K* and a likelihood ratio test for lambda. We tested for phylogenetic signal in qualitative (coded as binary) traits using Fritz and Purvis' *D* [@Fritz2010] with the \function{phylo.d} function in the R package \package{caper} [@Orme2018], which also assesses significance by comparing the observed value to a null distribution of random values and a simulated distribution of values expected under BM.

@@

We measured functional diversity (FD) following the method of @Petchey2002. We first calculated functional distances using Gower's method [@Gower1971] on the trait data with the \function{gowdis} function in the R package \package{FD} [@Laliberte2011]. We weighted continuous and qualitative traits equally, weighted each binary component trait of each qualitative trait equally, and log-transformed and scaled continuous traits prior to calculating distances. We then used the distances to build a trait dendrogram using the \function{hclust} function in the \package{stats} R package [@RCoreTeam2020], and used the dendogram in place of a phylogenetic tree to calculate PD and RPD as described above. These functional analogs of PD and RPD are hereafter referred to as FD (functional diversity) and RFD (relative functional diversity).

@@

Observed values of PD, FD, and PE are known to be closely related to species richness: as more taxa are drawn at random without replacement from a given set, it becomes increasingly unlikely to draw a taxon that is distantly related (or functionally dissimilar) to the others. Therefore, we determined whether the observed metrics are more or less diverse than random for a given number of taxa by conducting a randomization test. We generated 999 random communities using the "independent swap" algorithm of @Gotelli2000, which conserves richness per site and species occurrence frequencies while randomizing species identities, then compared the observed value to the distribution of random values. Statistical significance was assessed using a two-tailed test with alpha = 0.5; observed values > 97.5% or < 0.25% were considered significant.

We tested whether endemic areas have an over-representation of old (paleoendemic) or new (neoendemic) lineages using categorical analysis of neo- and paleo-endemism (CANAPE), which involves measuring PE with an alternate tree where all branch lengths are equal, then comparing the ratio between raw PE and alternative PE (RPE) [@Mishler2014]. Low RPE indicates short branchlengths (neoendemism) and high RPE indicates long branchlengths (paleoendemism). Since all of these measures are affected by species richness, categorization of endemism is done using the rank of each observed variable relative to those calculated for the null distribution. We used the same set of null communities as for SES of PE, and carried out categorization using custom R scripts.

@@

We downloaded SHP files of protected areas in Japan from the Japan Ministry of the Environment (https://www.biodic.go.jp/biodiversity/activity/policy/map/map17/index.html). We only considered areas that include protection of plants, and categorized protected areas as either "high" (no human activities allowed at all) or "medium" (some human activities allowed by permit). We cropped grid cells to only include land area, then calculated the percent of protected area of each status type within grid cells with high levels of biodiversity as measured with PD, FD, PE, or taxonomic richness. For PD, FD, and PE, significance was assessed with a one-tailed test at alpha = 0.05 against the 999 null communities described above. For taxonomic richness, cells ranked in the top >5% were considered highly diverse.

@@

```{r dataset-stats}
### Phylogeny ###

# FIXME: confirm total number of taxa with Ebihara
total_fern_taxa <- 675
phylogeny_sampled_percent <- ape::Ntip(japan_fern_tree) %>% magrittr::divide_by(total_fern_taxa) %>% scales::percent(accuracy = 0.01)

# Check height (age) of fern phylogeny
phylogeny_age <- japan_fern_tree %>% phytools::nodeHeights() %>% max %>% round(1)

# Check monophyly of genera
mono_status_genera <-
tibble(
  tip = japan_fern_tree$tip.label,
  tip_no = 1:length(tip),
  genus = str_split(tip, "_") %>% map_chr(1)
) %>%
  group_by(genus) %>%
  summarize(tip_vec = list(tip_no), .groups = "drop") %>%
  mutate(is_mono = map_lgl(tip_vec, ~ape::is.monophyletic(phy = japan_fern_tree, tips = .)))

# Assert only one genus is not monophyletic
assert_that(mono_status_genera %>% filter(is_mono == FALSE) %>% nrow == 1)

non_mono_genus <- mono_status_genera %>% filter(is_mono == FALSE) %>% pull(genus)

n_fern_phylo_genera <- mono_status_genera %>% pull(genus) %>% n_distinct()

# Check monophyly of families
mono_status_families <-
tibble(
  tip = japan_fern_tree$tip.label,
  tip_no = 1:length(tip),
  genus = str_split(tip, "_") %>% map_chr(1)
) %>%
  left_join(ppgi, by = "genus") %>%
  group_by(family) %>%
  summarize(tip_vec = list(tip_no), .groups = "drop") %>%
  mutate(is_mono = map_lgl(tip_vec, ~ape::is.monophyletic(phy = japan_fern_tree, tips = .)))

# Assert all families are monophyletic
assert_that(mono_status_families %>% filter(is_mono == FALSE) %>% nrow == 0)

n_fern_phylo_families <- mono_status_families %>% pull(family) %>% n_distinct()

# Check monophyly of ferns
mono_status_class <- 
tibble(
  tip = japan_pterido_tree$tip.label,
  tip_no = 1:length(tip),
  genus = str_split(tip, "_") %>% map_chr(1)
) %>%
  left_join(ppgi, by = "genus") %>%
  group_by(class) %>%
  summarize(tip_vec = list(tip_no), .groups = "drop") %>%
  mutate(is_mono = map_lgl(tip_vec, ~ape::is.monophyletic(phy = japan_pterido_tree, tips = .)))

# Assert all classes (ferns, lycophytes) are monophyletic
assert_that(mono_status_class %>% filter(is_mono == FALSE) %>% nrow == 0)

### Specimens ###
n_specimens <- nrow(occ_point_data_ferns) %>% scales::number(big.mark = ",")
n_specimens_taxa <- occ_point_data_ferns %>% pull(taxon) %>% n_distinct()
n_specimens_taxa_percent <- n_specimens_taxa %>% magrittr::divide_by(total_fern_taxa) %>% scales::percent(accuracy = 0.01)

# Calculate mean redundancy
mean_redundancy <-
biodiv_ferns_spatial %>%
  select(redundancy) %>%
  st_set_geometry(NULL) %>%
  summarize(
    mean = mean(redundancy),
    sd = sd(redundancy)
  ) %>%
  mutate(text = glue::glue("{round(mean, 2)} ± {round(sd, 2)}")) %>%
  pull(text)

### Traits ###

# Count number of traits of each type in final dataset
n_binary_traits <- traits_summary %>% filter(trait_type == "binary") %>% nrow() %>% jntools::english2()
n_cont_traits <- traits_summary %>% filter(trait_type == "continuous") %>% nrow() %>% jntools::english2()
n_qual_traits <- traits_summary %>% filter(trait_type == "qualitative") %>% nrow() %>% jntools::english2()
n_qual_traits_as_binary <- traits_summary %>% filter(trait_type == "qualitative") %>% pull(n) %>% sum %>% jntools::english2()

# Calculate number/percent of missing data in trait matrix
fern_traits_missing_summary <-
fern_traits %>%
  pivot_longer(-taxon) %>%
  mutate(missing_data = is.na(value)) %>%
  summarize(
    total = n(),
    missing = sum(missing_data)
  )

percent_traits_missing <-
fern_traits_missing_summary$missing %>% 
  divide_by(fern_traits_missing_summary$total) %>%
  scales::percent(accuracy = 0.01)
```

The phylogeny included `r ape::Ntip(japan_fern_tree)` taxa (`r phylogeny_sampled_percent`) representing `r n_fern_phylo_genera` genera and `r n_fern_phylo_families` families of native Japanese ferns. The topology was in general agreement with other recent, large plastid fern phylogenies [@Schuettpelz2007; @Lehtonen2011; @Testo2016a]. Ferns were recovered as monophyletic, as were all families and genera except for *`r non_mono_genus`* (families follow @PteridophytePhylogenyGroupI2016). The age of the fern clade is estimated to be `r phylogeny_age` ma.

The cleaned list of specimens included `r n_specimens` specimens representing `r n_specimens_taxa` taxa (`r n_specimens_taxa_percent`). The distribution of collection locations was highly skewed, with > 3,000 specimens per grid cell northwest of Tokyo in Saitama prefecture, and many fewer elsewhere (`r s_figure("abundance")`A). Despite this, sampling redundancy was generally good [> 0.3; @Gonzalez-Orozco2014b] throughout the country (mean `r mean_redundancy`; errors are SD unless mentioned otherwise) (`r s_figure("abundance")`B). The final community data matrix included `r ncol(comm_ferns)` taxa and `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid cells. __ grid cells were excluded from the analysis due to low redundancy (< 0.1); these were mostly from Hokkaido and coastal areas.

After removing correlated and low-variance traits, the final trait set included `r n_binary_traits` binary trait, `r n_cont_traits` continuous traits, and `r n_qual_traits` qualitative traits (coded as `r n_qual_traits_as_binary` binary traits) (`r table("traits-used")`). The trait matrix included `r nrow(fern_traits)` taxa and `r fern_traits %>% select(-taxon) %>% ncol()` traits, with `r percent_traits_missing` missing data.

@@

```{r phy-sig-stats}
# Verify that min lambda >0.95
assert_that(phy_sig_results %>% pull(lambda) %>% min > 0.95)
# Verify that max K < 0.0.6
assert_that(phy_sig_results %>% pull(kval) %>% max < 0.06)

# Sum the number of traits with D < 0
n_negative_d <-
binary_sig_results %>%
  mutate(negative_d = ifelse(D <= 0, TRUE, FALSE)) %>%
  pull(negative_d) %>%
  sum
```

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.06, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`). `r n_negative_d` out of `r nrow(binary_sig_results)` of continuous traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (*D* =< 0); others had *D* > 0 (`r s_table("phy-sig-binary")`).

@@

The latitudinal hump-shaped pattern of species richness observed in ferns here (`r figure("raw-diversity")`A) is well-documented [@Kusumoto2017; @Ebihara2019b], and similar patterns have been observed in deciduous broadleaved trees and conifers [@Shiono2015] and modeled distributions of woody plants [@Fukaya2020]. This pattern is probably due at least in part to the distribution of land area in Japan: maximum land area is located at mid-latitudes, and area is related to species richness by the well-known species-area relationship [@Arrhenius1921]. There may also be a combined climatic and elevational effect: the southern part of Honshu is warmer and more humid than areas further north, while also harboring high elevational variation that may support a wide range of taxa, which is not present in the smaller islands further south.

@@

<!-- Tables -->

```{r traits-used, results = "asis"}
trait_signif <- read_csv(here::here("data_raw/trait_functions.csv"))

traits_summary %>%
  left_join(trait_signif, by = "trait") %>%
  mutate(
    trait = case_when(
      trait == "frond_width" ~ "Frond width",
      trait == "number_pinna_pairs" ~ "No. pinna pairs",
      trait == "stipe_length" ~ "Stipe length",
      trait == "margin" ~ "Frond margin",
      trait == "shape" ~ "Frond shape",
      trait == "texture" ~ "Frond texture",
      TRUE ~ str_to_sentence(trait)
    ),
    trait_type = case_when(
      trait_type == "qualitative" ~ glue::glue("Qualitative (coded as {n} binary traits)"),
      trait_type == "continuous" & str_detect(trait, "pinna pairs") ~ glue::glue("Continuous (integer)"),
      trait_type == "continuous" ~ glue::glue("Continuous (cm)"),
      TRUE ~ str_to_sentence(trait_type)
    )
  ) %>%
  select(trait, Significance = significance) %>%
  column_to_rownames("trait") %>%
knitr::kable(format = "latex", booktabs = TRUE, escape = TRUE)

**`r table("traits-used")`**. `r table_cap("traits-used")`
